---
title: "The Linear Model"
---


The linear model is the foundation of statistical hypothesis testing.
Many of the analyses we encounter in the technical literature can be seen as extensions of the linear model.
Understanding experimental design requires an understanding of what test we plan to perform.
A foundation in the linear model will help us understand the design and implementation of our experiments.


## The linear equation


In grade school geometry we learned the equation of a line.

$$
y = mx + b
$$


Here, `x` and `y` are coordinates on the plane.
The slope is `m`.
And the y-intercept is `b`.
We can plot this in R as follows.


```{r, fig.align='center', fig.cap='Figure 1. A straight line.'}
x <- 1:20
y <- 0.5 * x + 2
plot(x, y)
```


The first line generates some data and assigns it to the variable `x`.
The second line uses a contrived slope (0.5) and intercept (2) to generate values which we assign to the variable `y`.
Because the values of `y` **depend** on `x` we sometimes call `y` the **dependent** variable and `x` the **independent** variable.
Similarly, we may refer to `y` as the **response** and `x` as **predictor(s)**.
The line crosses the y-intercept at one and for each increase of value in `x` we see a half increase in `y`.


## Adding the error term


In biology we rarely see such clear patterns.
In order to accomodate this variation we add an error term.

$$
y = mx + b + \epsilon_{Norm}
$$


This can be plotted in R as follows.


```{r, fig.align='center', fig.cap='Figure 2. A straight line with random error.'}
set.seed(9)
y <- 0.5 * x + 2 + rnorm(n=length(x), mean = 0, sd = 1)
plot(x, y)
```


This is the exact same plot as in Figure 1 except that the data are now distributed around the line with normally distributed error.
The mean of this distribution was zero because we do not want it to deviate from the line.
The standard deviation, or the dispersal around the line, was arbitrarily set to one.
In your experiments the standard deviation will represent biological variation combined with technical error assotiated with measuring the data.

### Linear regression

When we use linear regression we are attempting to infer the slope and intercept that we've specified above.
We can use the function `lm()` to perform the regression and it's `summary()` method to report the results.


```{r}
lm1 <- lm(y ~ x)
summary(lm1)
```


We see that the intercept is ```r lm1$coefficients[1]```, which should be close to two, and the slope is ```r lm1$coefficients[2]```, which should be close to 0.5.


## Analysis of variance


In the analysis of soilborne pathogens we are typically looking for differences among groups.
In linear regression we were looking for relationships among a (putatively) dependent variable and independent variables.



